Wide binaries in Gaia DR1

### 2016-09-29 paper/ms.tex
fix signs, other bugs, tweaks

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{N}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\data}{\mathrm{data}}

@@

For a pair of stars $(i,j)$, we want to ratio the fully marginalized likelihood
(FML) of the stars having the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) over the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). When this ratio is large, the pair is a
candidate binary star. For \gaia\ DR1, we only observe parallax, $\varpi$, and
two proper motion components, $\propm = (\mu_\alpha, \mu_\delta)$ for each star.
Where there is overlap with \rave\ or \apogee, some number of observed radial
velocities, $v_r$, with observed variances $\sigma^2_{v_r}$ (independent of the
\gaia\ data) may also be available. We assume that the uncertainties in these
observed quantites are known with covariance matrix $\matrx{C}$. The likelihoods
$\mathcal{L}_1, \mathcal{L}_2$ are marginalized over true 3-space velocity and
distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i) \, p(d_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i) \, p(d_j) \\
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\begin{array}{c c c c} \varpi & \mu_\alpha &
    \mu_\delta & v_r \end{array}\right)} \\
  \bs{x}_\theta &= \transp{\left(\begin{array}{c c c c} \frac{1}{d} & \frac{v_\alpha}{d} &
    \frac{v_\delta}{d} & \tilde{v}_r \end{array}\right)}
\end{align}
\todo{Holy fuck that vector is UGLY}
We assume Gaussian priors on the true velocity components and use a \todo{which
one will we use?} distance prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{align}
  p(\bs{v}) &= \mathcal{N}(0,25)~\kms \\
  p(d) &= \frac{r^2}{2\,L^3} \, \exp\left(-r/L\right)
\end{align}

@@

In what follows, all vectors are column vectors.
We indicate whether a symbol represents a scalar, vector, or tensor
(matrix) by context, not typesetting.
Properties of the squared exponential:
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] - \Delta)\,\dd x\right]
  &=& \frac{D}{2}\,\ln 2\pi +\frac{1}{2}\ln ||A|| -\Delta
  \quad ,
\end{eqnarray}
where $x$ and $\nu$ are $D$-dimensional vectors, $A$ is a symmetric
matrix, $\Delta$ is a scalar, and the integral is implicitly over all
of $D$-dimensional $x$-space.
We will need to complete squares.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] + \Delta
  &=& \frac{1}{2}\,\transp{x}\,\inv{A}\,x + \transp{x}\,B\,b + C
  \quad ,
\end{eqnarray}
where $B\,b$ is a $d$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \nu &=& A\,B\,b
  \\
  C & = & \Delta - \frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad .
\end{eqnarray}

Now in this case, we want to write down the probability of the data
given the distance and velocity (the likelihood), and the prior pdf for
the velocity, and their product.
The prior pdf for the velocity is Gaussian:
\begin{eqnarray}
  p(v) &=& \normal(v\given 0,V)
  \\
  \ln p(v) &=& -\frac{3}{2}\,\ln 2\pi -\frac{1}{2}\,\ln||V|| -\frac{1}{2}\,\transp{v}\,\inv{V}\,v
  \\
  V &=& \sigma_v^2\,I
  \quad ,
\end{eqnarray}
where $v$ is the three-vector velocity, and $V$ is an isotropic
variance tensor.
Similarly, the likelihood for the data is also a Gaussian.
Before we write that down, we will make some definitions.
We construct a vector $y$ as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d_1\,\varpi_1 - 1), d_1\,\mu_{\alpha,1}, d_1\,\mu_{\delta,1},
                       d_1\,\varpi_2 - 1, d_2\,\mu_{\alpha,2}, d_1\,\mu_{\delta,2}]
  \\
  \transp{y} &\equiv& [(d_1\,\varpi_1 - 1), d_1\,\mu_{\alpha,1}, d_1\,\mu_{\delta,1}, v_{r,1}
                       d_1\,\varpi_2 - 1, d_2\,\mu_{\alpha,2}, d_1\,\mu_{\delta,2}, v_{r,2}
  \quad ,
\end{eqnarray}
where we have given two versions, a 6-component and an 8-component
version (depending on whether there are measured radial velocities),
the subscripts 1, 2 refer to the first and second stars (arbitrarily
ordered) of the pair, we have multiplied the strict observables by
distances $d_1, d_2$, which we are permitted to do (because we are
\emph{conditioning} on $d_1, d_2$), and note the wacky minus-ones.
The vector $y$ is a column vector that we have transposed to make it
easy to write down.
Fundamentally, our model is
\begin{eqnarray}
  y &=& M\,v + \mathrm{noise}
  \\
  \transp{M} &=& [0, \hat{\alpha_1}, \hat{\delta_1},
                  0, \hat{\alpha_2}, \hat{\delta_2}]
  \\
  \transp{M} &=& [0, \hat{\alpha}_1, \hat{\delta}_2, \hat{r}_1,
                  0, \hat{\alpha}_2, \hat{\delta}_2, \hat{r}_2]
  \quad,
\end{eqnarray}
where $M$ is a $6\times 3$ or $8\times 3$ design matrix, with zeros
in two rows, and unit vectors along the other rows.
The noise (in the $y$ vector) is drawn from a Gaussian with ($6\times 6$ or $8\times 8$)
covariance matrix $\Sigma$, constructed from the distances $d_1, d_2$ and
the \gaia-provided individual-object covariance matrices.
Given all these definitions, the likelihood function is
\begin{eqnarray}
  p(\data\given v,d_1,d_2) &=& d_1^3\,d_2^3\,\normal(y\given M\,v, \Sigma)
  \\
  \ln p(\data\given v,d_1,d_2) &=& 3\,\ln d_1 + 3\,\ln d_2 -\frac{6~\mathrm{or}~8}{2}\,\ln 2\pi
  -\frac{1}{2}\,\ln||\Sigma|| \\ && \quad -\frac{1}{2}\,\transp{[y-M\,v]}\,\inv{\Sigma}\,[y-M\,v]
  \quad ,
\end{eqnarray}
where the factor of $d_1^2\,d_2^3$ converts to units of \gaia\ data
from units of $y$.

Now we multiply together (add in the log) and complete the square.
We get, in the original parameterization:
\begin{eqnarray}
  \nu &=& \inv{[\transp{M}\,\inv{\Sigma}\,M+\inv{V}]}\,\transp{M}\,\inv{\Sigma}\,y
  \\
  \Delta &=& 3\,\ln d_1 + 3\,\ln d_2 -\frac{9~\mathrm{or}~11}{2}\,\ln 2\pi
  -\frac{1}{2}\,\ln||\Sigma|| -\frac{1}{2}\,\ln||V|| \\
  &&\quad -\frac{1}{2}\,\transp{y}\,\inv{\Sigma}\,y -\frac{1}{2}\,\transp{\nu}\,\inv{[\transp{M}\,\inv{\Sigma}\,M+\inv{V}]}\,\nu
  \\
  \ln p(\data\given d_1,d_2) &=& [HOGG]
  \quad ,
\end{eqnarray}
where....HOGG

>>>>

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\data}{\mathrm{data}}

@@

For a pair of stars $(i,j)$, we want to ratio the fully marginalized likelihood
(FML) of the stars having the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) over the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). When this ratio is large, the pair is a
candidate binary star. For \gaia\ DR1, we only observe parallax, $\varpi$, and
two proper motion components, $\propm = (\mu_\alpha, \mu_\delta)$ for each star.
Where there is overlap with \rave\ or \apogee, some number of observed radial
velocities, $v_r$, with observed variances $\sigma^2_{v_r}$ (independent of the
\gaia\ data) may also be available. We assume that the uncertainties in these
observed quantites are known with covariance matrix $\matrx{C}$. The likelihoods
$\mathcal{L}_1, \mathcal{L}_2$ are marginalized over true 3-space velocity and
distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i) \, p(d_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i) \, p(d_j) \\
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\begin{array}{c c c c} \varpi & \mu_\alpha &
    \mu_\delta & v_r \end{array}\right)} \\
  \bs{x}_\theta &= \transp{\left(\begin{array}{c c c c} \frac{1}{d} & \frac{v_\alpha}{d} &
    \frac{v_\delta}{d} & \tilde{v}_r \end{array}\right)}
\end{align}
\todo{Holy fuck that vector is UGLY}
We assume Gaussian priors on the true velocity components and use a \todo{which
one will we use?} distance prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{align}
  p(\bs{v}) &= \normal(0,25)~\kms \\
  p(d) &= \frac{r^2}{2\,L^3} \, \exp\left(-r/L\right)
\end{align}

@@

In what follows, all vectors are column vectors.
We indicate whether a symbol represents a scalar, vector, or tensor
(matrix) by context, not typesetting.
Properties of the squared exponential:
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] - \Delta)\,\dd x\right]
  &=& \frac{D}{2}\,\ln 2\pi +\frac{1}{2}\ln ||A|| -\Delta
  \quad ,
\end{eqnarray}
where $x$ and $\nu$ are $D$-dimensional vectors, $A$ is a positive definite
matrix, $\Delta$ is a scalar, and the integral is implicitly over all
of $D$-dimensional $x$-space.
We will need to complete squares.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[x-\nu]}\,\inv{A}\,[x-\nu] + \Delta
  &=& \frac{1}{2}\,\transp{x}\,\inv{A}\,x + \transp{x}\,B\,b + C
  \quad ,
\end{eqnarray}
where $B\,b$ is a $d$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \nu &=& A\,B\,b
  \\
  \Delta & = & C - \frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad .
\end{eqnarray}

Now in this case, we want to write down the probability density for the data
given the distances and velocity (the likelihood), and the prior pdf for
the velocity, and their product.
The prior pdf for the velocity is Gaussian:
\begin{eqnarray}
  p(v) &=& \normal(v\given 0,V)
  \\
  \ln p(v) &=& -\frac{3}{2}\,\ln 2\pi -\frac{1}{2}\,\ln||V|| -\frac{1}{2}\,\transp{v}\,\inv{V}\,v
  \\
  V &=& \sigma_v^2\,I
  \quad ,
\end{eqnarray}
where $v$ is the three-vector velocity, and $V$ is an isotropic
variance tensor.
Similarly, the likelihood for the data is also a Gaussian.
Before we write that down, we will make some definitions.
We construct a vector $y$ as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [(d_1\,\varpi_1 - 1), d_1\,\mu_{\alpha,1}, d_1\,\mu_{\delta,1},
                       d_1\,\varpi_2 - 1, d_2\,\mu_{\alpha,2}, d_1\,\mu_{\delta,2}]
  \\
  \transp{y} &\equiv& [(d_1\,\varpi_1 - 1), d_1\,\mu_{\alpha,1}, d_1\,\mu_{\delta,1}, v_{r,1}
                       d_1\,\varpi_2 - 1, d_2\,\mu_{\alpha,2}, d_1\,\mu_{\delta,2}, v_{r,2}
  \quad ,
\end{eqnarray}
where we have given two versions, a 6-component and an 8-component
version (depending on whether there are measured radial velocities),
the subscripts 1, 2 refer to the first and second stars (arbitrarily
ordered) of the pair, we have multiplied the strict observables by
distances $d_1, d_2$, which we are permitted to do (because we are
\emph{conditioning} on $d_1, d_2$), and note the wacky minus-ones.
The vector $y$ is a column vector that we have transposed to make it
easy to write down.
Fundamentally, our model is
\begin{eqnarray}
  y &=& M\,v + \mathrm{noise}
  \\
  \transp{M} &=& [0, \hat{\alpha_1}, \hat{\delta_1},
                  0, \hat{\alpha_2}, \hat{\delta_2}]
  \\
  \transp{M} &=& [0, \hat{\alpha}_1, \hat{\delta}_2, \hat{r}_1,
                  0, \hat{\alpha}_2, \hat{\delta}_2, \hat{r}_2]
  \quad,
\end{eqnarray}
where $M$ is a $6\times 3$ or $8\times 3$ design matrix, with zeros
in two rows, and unit vectors along the other rows.
The noise (in the $y$ vector) is drawn from a Gaussian with ($6\times 6$ or $8\times 8$)
covariance matrix $\Sigma$, constructed from the distances $d_1, d_2$ and
the \gaia-provided individual-object covariance matrices.
Given all these definitions, the likelihood function is
\begin{eqnarray}
  p(\data\given v,d_1,d_2) &=& d_1^3\,d_2^3\,\normal(y\given M\,v, \Sigma)
  \\
  \ln p(\data\given v,d_1,d_2) &=& 3\,\ln d_1 + 3\,\ln d_2 -\frac{6~\mathrm{or}~8}{2}\,\ln 2\pi
  -\frac{1}{2}\,\ln||\Sigma|| \\ && \quad -\frac{1}{2}\,\transp{[y-M\,v]}\,\inv{\Sigma}\,[y-M\,v]
  \quad ,
\end{eqnarray}
where the factor of $d_1^3\,d_2^3$ converts to units of per \gaia\ data
from units of per $y$.

Now we multiply together (add in the log) and complete the square.
We get, in the original parameterization:
\begin{eqnarray}
  \inv{A} &=& [\transp{M}\,\inv{\Sigma}\,M+\inv{V}]
  \\
  \nu &=& A\,\transp{M}\,\inv{\Sigma}\,y
  \\
  \Delta &=& -3\,\ln d_1 -3\,\ln d_2 +\frac{9~\mathrm{or}~11}{2}\,\ln 2\pi
  +\frac{1}{2}\,\ln||\Sigma|| +\frac{1}{2}\,\ln||V|| \\
  &&\quad +\frac{1}{2}\,\transp{y}\,\inv{\Sigma}\,y -\frac{1}{2}\,\transp{\nu}\,\inv{A}\,\nu
  \quad ,
\end{eqnarray}
where there are sign issues that Hogg doesn't like to think about.
Finally, we marginalize out the velocity to get the marginalized
likelihood conditioned on the two distances $d_1, d_2$:
\begin{eqnarray}
  \ln p(\data\given d_1,d_2)
  &=& \frac{3}{2}\,\ln 2\pi +\frac{1}{2}\ln ||A|| -\Delta
  \quad .
\end{eqnarray}
